---
title: Adventures in `httptest2`
subtitle: Testing API requests in R packages
date: 2025-02-02T17:41:38-05:00
draft: true
tags: []
categories: []
format: hugo-md
knitr:
    opts_chunk:
        fig.path: "images/"
author:
  name: "Ali Sajid Imami"
  link: "/about_me/"
  avatar: "/images/aliimami.png"
  email: "<hello@aliimami.com>"
---

```{r}
#| label: setup
#| echo: false
#| include: false


knitr::opts_chunk$set( # nolint: extraction_operator_linter.
    collapse = TRUE,
    comment = "#>",
    cache = FALSE,
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.path = "images/"
)

suppressPackageStartupMessages({
    library(httr2)
    library(httptest2)
})
```


# Introduction

This blog post is part rant, part catharsis, and part education. The emotions
are due to the circumstances in which I discovered the problem, while the 
eduction part comes from my desire to not have other people go through
what I did.

I maintain an R package called [`drugfindR`](https://www.github.com/CogDisResLab/drugfindR).
This package is important to me for several reasons, including but not limited to:

* It was the first __proper__ R package I ever wrote
* It came out of my experiences (and publications) during COVID-19 lockdowns
* It has been instrumental in [our lab's](https://cdrl-ut.org) drug repurposing
work
* It is part of my Dissertation

Part of the package's functionality relies on an external REST API and what
better way to interact with external APIs in R than using the [httr2](httr2.r-lib.org/)
package. Recently I have been working on modernizing the internal functionality
and refactoring giant functions into smaller, more focused functions. As I am 
creating more functions, I am also revamping my test suite. 

Previously, I had relied on the [vcr](https://docs.ropensci.org/vcr/index.html)
package to record API interactions and then use them in subsequent test runs.
This was combined with `httr2`'s convenience functions and a harness that 
I had built precisely for this purpose. This was sufficient for my purposes,
but it was still messy. I did not like that kind of mess in my codebase.

One day, as I was going through `httr2`'s documentation, looking for
yet another way to simplify my mock responses and requests, I came across this
little gem in the documentation for the [`with_mocked_responses()`] function:

> Instead use package that uses these functions under the hood, like httptest2 or vcr.

Thus, began my introduction to the [`httptest2`] package and the yak-shavingest
upgrade process for my tests.


# Introduction to `httptest2`


```{r}
hist(iris$Sepal.Width)
```


{{< center-quote >}}
So long, and thanks for all the fish.
{{< /center-quote >}}
